---
title: "FinalProject"
author: "Mai Le"
date: "5/17/2019"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(readxl)
library(dplyr) #grouping and manipulating data
library(ggplot2) #data visualizations
library(mice) #missing data imputation
```

```{r, echo=FALSE, include=FALSE}
mydata <- read_excel("~/Documents/SASUniversityEdition/myfolders/FinalProject/AchievementFirstDataset.xlsx")

#remove column Student ID 
mydata <- select(mydata, -1)

# check for invalid reading scores and convert them into missing values (NA)
mydata$BOY_Score <- ifelse(mydata$BOY_Score <1 | mydata$BOY_Score >26, NA, mydata$BOY_Score)  
mydata$EOY_Score <- ifelse(mydata$EOY_Score <1 | mydata$EOY_Score >26, NA, mydata$EOY_Score) 

mydata <- mutate(mydata, Score_diff = EOY_Score - BOY_Score)

#convert variables into factors
mydata$Grade <- factor(mydata$Grade)
mydata$School <- factor(mydata$School)

```
## Background and Executive Summary
```{r, echo=FALSE, warning==FALSE}
```
## Exploratory Data Analysis
## find average reading score by school and grade
Create a separate dataset for reading scores
``` {r}
# select only relevant variables
df2 <- data.frame(ID = edited$Student_ID,
                  School = edited$School,
                  Grade = edited$Grade,
                  BOY_score = edited$BOY_Score,
                  EOY_score = edited$EOY_Score)
# convert data into long format
df2.long <- reshape(df2, direction = "long", varying = list(names(df2)[4:5]), 
                    v.names = "Score", 
                    idvar = c("ID"),
                    timevar = "Period", times = c("BOY_score", "EOY_score"))
df2.long <- within(df2.long, {
  ID <-factor(ID)
  Period <-factor(Period, labels=c('Beginning-Of-Year', 'End-Of-Year'))
})

# a similar data frame under listwise deletion
#Listwise deletion
data.listwise <- mydata[complete.cases(mydata),]
# convert reading scores into proficiency levels
edited.listwise <- mutate(data.listwise, BOY_level = ifelse(Grade=='5' & BOY_Score %in% 1:9, "Remedial",
                                            ifelse(Grade=='5' & BOY_Score %in% 10:11, "Below_Proficient",
                                            ifelse(Grade=='5' & BOY_Score %in% 12:13, "Proficient",
                                            ifelse(Grade=='5' & BOY_Score >13, "Advanced",
                                            ifelse(Grade=='6' & BOY_Score %in% 1:11, "Remedial",
                                            ifelse(Grade=='6' & BOY_Score %in% 12:13, "Below_Proficient",
                                            ifelse(Grade=='6' & BOY_Score %in% 14:15, "Proficient",
                                            ifelse(Grade=='6' & BOY_Score >15, "Advanced", NA)))))))))       
edited.listwise <- mutate(edited.listwise, EOY_level = ifelse(Grade=='6' & EOY_Score %in% 1:13, "Remedial",
                                              ifelse(Grade=='6' & EOY_Score %in% 14:15, "Below_Proficient",
                                              ifelse(Grade=='6' & EOY_Score %in% 16:17, "Proficient",
                                              ifelse(Grade=='6' & EOY_Score >17, "Advanced",
                                              ifelse(Grade=='5' & EOY_Score %in% 1:11, "Remedial",
                                              ifelse(Grade=='5' & EOY_Score %in% 12:13, "Below_Proficient",
                                              ifelse(Grade=='5' & EOY_Score %in% 14:15, "Proficient",
                                              ifelse(Grade=='5' & EOY_Score >15, "Advanced", NA)))))))))

#convert BOY_level and EOY_level into factors
edited.listwise$BOY_level <- as.factor(edited.listwise$BOY_level)
edited.listwise$EOY_level <- as.factor(edited.listwise$EOY_level)
df2.lw <- data.frame(ID = edited.listwise$Student_ID,
                  School = edited.listwise$School,
                  Grade = edited.listwise$Grade,
                  BOY_score = edited.listwise$BOY_Score,
                  EOY_score = edited.listwise$EOY_Score)
df2.lw.long <- reshape(df2.lw, direction = "long", varying = list(names(df2)[4:5]), 
                    v.names = "Score", 
                    idvar = c("ID"),
                    timevar = "Period", times = c("BOY_score", "EOY_score"))
df2.lw.long <- within(df2.long, {
  ID <-factor(ID)
  Period <-factor(Period, labels=c('Beginning-Of-Year', 'End-Of-Year'))
})

## Find average reading scores for Bushwick
bushwick.raw <- filter(df2.long, School == 'Bushwick MS')
bushwick.summary.raw <- ddply(bushwick.raw, ~Grade*Period, summarise, mean=mean(Score, na.rm = T), sd=sd(Score, na.rm = T))
# results under listwise
bushwick.raw.lw <- filter(df2.lw.long, School == 'Bushwick MS')
bushwick.summary.raw.lw <- ddply(bushwick.raw.lw, ~Grade*Period, summarise, mean=mean(Score, na.rm = T))

## Find average reading scores for Crown Heights
crownheights.raw <- filter(df2.long, School == 'Crown Heights MS')
crownheights.summary.raw <- ddply(crownheights.raw, ~Grade*Period, summarise, mean=mean(Score, na.rm = T), sd=sd(Score, na.rm = T))
# results under listwise
crownheights.raw.lw <- filter(df2.lw.long, School == 'Crown Heights MS')
crownheights.summary.raw.lw <- ddply(crownheights.raw.lw, ~Grade*Period, summarise, mean=mean(Score, na.rm = T))

```{r, echo=TRUE, warning=FALSE}
# scatterplot showing interaction effect between Grade and BOY_Score in predicting EOY_Score 
qplot(x=BOY_Score, y=EOY_Score, color=Grade, data=mydata) + geom_smooth(method='lm')

# scatterplot showing interaction between School and BOY_Score in predicting EOY_Score
qplot(x=BOY_Score, y=EOY_Score, color=School, data=mydata) + geom_smooth(method='lm')

#Plot panels to show how School, Grade, and BOY_Score affect EOY_Score
qplot(x=BOY_Score, y=EOY_Score, color=Grade, facets=School ~ ., data=mydata)

#scatterplot showing Score improvement by Grade and School
qplot(x=BOY_Score, y=Score_diff, color=Grade, data=mydata) + geom_smooth(method='lm')

qplot(x=BOY_Score, y=Score_diff, color=School, data=mydata) + geom_smooth(method='lm')

```
## Linear Model
```{r, echo=FALSE, include=FALSE}
##Model building predicting Score diff
mod1 = lm(Score_diff ~ School + Grade + BOY_Score + School*Grade + School*BOY_Score + Grade*BOY_Score, data = mydata)
par(mfrow=c(2,2))

#Backward Selection
step(mod1, direction='backward')

#Selected model
diff.mod = lm(Score_diff ~ School + Grade + BOY_Score + School:Grade + 
          Grade:BOY_Score, data = mydata)
summary(diff.mod)
plot(diff.mod)

#Check for influential points
cookd = cooks.distance(diff.mod)
plot(cookd)
```



